ActiveCA: Time Use Data from the General Social Survey of Canada to Study Active Travel

### 2024-12-04 paper/Manuscript-Data-Package.tex
Started editing the paper after Bruno's work

\begin{abstract}
This paper describes the open data product \texttt{ActiveCA}, with
Canadian time use data about active travel. \texttt{ActiveCA} is an
\texttt{R} data package that contains analysis-ready data from cycles
General Social Survey cicles regarding active travel in Canada. The
package provides data set detailing active trip origins, destinations,
and duration, covering a wide variety of locations, such as home, work
or school, libraries, museums, restaurants, bars, sports centers, health
clinics, place of worship, and others. The package also details the
respondent's region characteristics, specifying wheter they live in a
metropolitan area and their province of residency. \texttt{ActiveCA}
also provides pre-estimated impedance functions for active travel for
Canadian Metropolitan Area and Census Agglomerations, which can be used
in accessibility analysis to calculate the cost of travel between
different locations. The package will continue to expand with
contributions from the authors and the broader community through
requests in the future. \texttt{ActiveCA} is freely accessible for
exploration and download from the associated Github repository, where
the documentation and code involved in creating and manipulating data
are detailed.
\end{abstract}

\keywords{Active; mobility; walking; cycling; travel time; impedance;
transportation; R}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

This paper presents the open data product \texttt{ActiveCA}. Open data
products (ODPs) are the outcome of a transparent process that transforms
raw data (open or not) into analysis-ready data, in which all stages of
development follow open principles \citep{arribas-bel2021}. ODPs differ
from general open data due to their high utility, added value and open
availability.

\texttt{ActiveCA} is an \texttt{R} data package that provides
analysis-ready data from the General Social Survey of Canada Cycles 2
(1986), 7 (1992), 12 (1998), 19 (2005), 24 (2010), and 29 (2015)
\citep{statisticscanada2024} regarding active travel. These data can be
used to study walking and cycling in Canada for various purposes over
time, and to calculate impedance functions for accessibility analysis.
The package provides pre-estimated impedance functions for cycling and
walking trips for Census Agglomerations and Census Metropolitan Areas.

To create this \texttt{R} data package, we collected, cleaned and
processed the Time Use collections from the GSS surveys to make them
ready for analysis. An \texttt{R} data package contains code, data and
documentation in a standardized format that can be installed by
\texttt{R} users via a centralized software repository, such as CRAN
(Comprehensive R Archive Network) and GitHub. Although the GSS surveys
are publicly sourced managed by Statistic Canada
\citep{statisticscanada2024}, preparing them for analysis can be
time-consuming, tedious and perhaps not even possible for those who try,
due to a lack of documentation or prior knowledge.

The aim of this paper is to walk readers through the data sets and
invite others to experiment in its uses and applications.
\texttt{ActiveCA} is freely available on GitHub for all to install and
freely use in the spirit of open and reproducible research. In addition
of enabling to obtain impedance functions, \texttt{ActiveCA} can be
adopted in various applications that even go beyond the range of
possibilities we have imagined. Not only the data, but also all the code
documenting the processing methodology is available for consultation and
evaluation in its repository. This package contributes to reducing the
barrier to using the information contained in GSS surveys to provide
data-driven decisions in transportation analysis.

\hypertarget{general-social-survey-gss-collection}{%
\section{General Social Survey (GSS)
collection}\label{general-social-survey-gss-collection}}

Statistics Canada \citeyearpar{statisticscanada2024} conducts GSS
surveys to obtain data on social trends to track changes in Canadians'
living conditions and well-being over time. This survey is used to
understand how citizens spend and manage their time and what factors
contribute to their happiness and stress. Created in 1985, the survey is
part of a series of independent, annual, and cross-sectional surveys.

In addition to the main topic, each GSS cycle includes new content that
addresses emerging and policy-relevant issues. Every five to seven
years, the Time Use Surveys \citep{statisticscanada2022} collect data on
respondents' participation and time spent on a wide range of everyday
activities using a 24-hour retrospective diary, with information on the
location of these activities (e.g.~at home, at work, etc.) and, for
non-personal activities, the people who were present with the respondent
at the time of the activity. In addition, time-use surveys also cover
topics related to leisure time, work-life balance, health, commuting,
culture and sports, and many others.

The most recent time use survey was carried out in 2022
\citep{wray2024}. However, the 2022 dataset has not been fully published
and, because of this, our analysis focused on the surveys from 1986 to
2015 (1986, 1992, 1998, 2005, 2010 and 2015). Time Use surveys are
composed of two data sets, the main one and the episode file, explained
in the following subsections.

\hypertarget{the-main-file}{%
\subsection{The Main File}\label{the-main-file}}

The main file compiles an large array of aggregated data, summarizing
the answers to the questionnaire and derived variables that summarize
the respondents' time use across different activities, locations, and
social interactions. This file documents the time and duration that
respondents allocate to each activity and location. The Main File
provides a overview of daily routines and social dynamics, not focusing
on individual activity episodes. Additionally, this file categorizes
activities into bigger groups and subcategories, facilitating the data's
analytical utility with additional metrics such as total transit time,
duration spent with household members, and counts of activities and
episodes.

\hypertarget{the-episode-file}{%
\subsection{The Episode File}\label{the-episode-file}}

The episode file records detailed data for each activity episode
reported by respondents. The entry includes the start and end times,
duration, location, and accompanying social context, informing when and
where activities occurred and with whom. The file distinguishes itself
by focusing on individual episodes rather than respondents, with the
data structured around the numerous activity instances that compose a
day of the respondent. Although respondent-specific characteristics are
not included within the episode file, it is possible to link the main
file and the episode file by using an identifiable variable present in
both data sets.

\hypertarget{methodology}{%
\section{Methodology}\label{methodology}}

For each selected cycle of the GSS surveys, we reviewed the episode
files to identify cases with activities listed as walking or cycling,
selecting the activities immediately before and after the mobility
episode. After that, we labeled the code variables with their
appropriate descriptions, identifying each origin and destination,
transportation mode, as well the province and urban classification of
the respondent's residency.

Additionally, for Census Agglomerations and Census Metropolitan Areas we
pre-estimated impedance functions for cycling and walking trips.
Impedance functions reveals important information about the travel
behavior of the population, by describing the relationship between the
population at an origin and their likelihood or ability to travel to
specific destinations to access opportunities \citep{soukhov2024}.

Impedance functions are commonly used in accessibility analysis to
calculate the cost of travel between different locations
\citep{hansen1959, páez2012, palacios2022}. However, these functions
need to be calibrated in order to accurately represent travel behavior.
One effective way of calibrating an impedance function is by using the
trip length distribution (TLD) obtained from origin-destination data
\citep{soukhov2024}. The TLD represents the probability that a certain
proportion of trips will be made at a specific cost, such as travel
time. In this data set, low travel times are associated with a higher
proportion of trips, while high travel times are associated with a lower
proportion of trips.

For each combination of year, destination, and transportation mode
(walking or cycling), we fitted the most suitable impedance function
based on empirical data from the GSS surveys. We used the
\texttt{fitdistrplus} package \citep{mullerdutang2015} to estimate the
functions, selecting the distribution with the lowest Akaike information
criterion (AIC) among exponential, gamma, log-normal, normal, and
uniform types.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{descriptive-statistics}{%
\subsection{Descriptive statistics}\label{descriptive-statistics}}

@@

\texttt{ActiveCA} also enables visual analysis of active travel in
Canada using traditional exploratory data analysis techniques. Figures
\ref{fig:figure-01} and \ref{fig:figure-02} show walking and cycling
trips from 1992 and 2015 through heat maps. These maps use color
gradients to represent the percentage of trips between various origins
and destinations, with darker colors indicating higher percentages and
lighter colors representing less frequent routes. To avoid overwhelming
the reader, we omitted the heat maps for the other years analyzed.

@@

\hypertarget{pre-estimated-impedance-functions-for-active-travel-in-canada}{%
\subsection{Pre-estimated impedance functions for active travel in
Canada}\label{pre-estimated-impedance-functions-for-active-travel-in-canada}}

\texttt{ActiveCA} includes a total of 64 pre-estimated impedance
functions for cycling and walking trips for Census Agglomerations and
Census Metropolitan Areas. Table \ref{tab:table-04} shows the best
impedance functions for walking trips to \texttt{Work\ or\ school}
across survey years, where all distributions, except for a gamma
function in 1998, are log-normal.

\begin{table}
\centering
\caption{\label{tab:table-04}\label{table-04}Impedance functions and AIC for `Walking` trips considering `Work or school` as destination.}
\caption{\label{tab:table-04}\label{table-04}Impedance functions and AIC for `Cycling` trips considering `Work or school` as destination.}
\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\fontsize{10}{12}\selectfont
\begin{threeparttable}
\begin{tabular}[t]{lrlrrrr}
\toprule
\multicolumn{1}{c}{\textbf{Mode}} & \multicolumn{1}{c}{\textbf{Year}} & \multicolumn{1}{c}{\textbf{Impedance function}} & \multicolumn{1}{c}{\textbf{Parameter 1*}} & \multicolumn{1}{c}{\textbf{Parameter 2*}} & \multicolumn{1}{c}{\textbf{AIC}} & \multicolumn{1}{c}{\textbf{Count}}\\
\midrule
 & 1992 & Gamma & 3.00 & 0.17 & 433582 & 19\\

 & 1998 & Gamma & 3.37 & 0.10 & 481536 & 19\\

 & 2005 & Lognormal & 2.93 & 0.70 & 888655 & 64\\

 & 2010 & Lognormal & 2.65 & 0.77 & 1292760 & 53\\

\multirow[t]{-5}{*}{\raggedright\arraybackslash Cycling} & 2015 & Lognormal & 3.03 & 0.41 & 1162876 & 63\\
\cmidrule{1-7}
 & 1992 & Lognormal & 2.38 & 0.70 & 2319400 & 113\\

 & 1998 & Gamma & 1.23 & 0.09 & 2318752 & 109\\

 & 2005 & Lognormal & 2.13 & 0.79 & 8182691 & 724\\

 & 2010 & Lognormal & 2.21 & 0.78 & 7917431 & 494\\

\multirow[t]{-5}{*}{\raggedright\arraybackslash Walking} & 2015 & Lognormal & 2.55 & 0.64 & 6612061 & 407\\
\bottomrule
\end{tabular}
\begin{tablenotes}
\item \textit{Note: } 
\item *For lognormal distributions, 'Parameter 1' and 'Parameter 2' refer to the mean and standard deviation of the distribution on the logarithmic scaler, respectively. For gamma distribution, 'Parameter 1' and 'Parameter 2' refer to the rate and shape of the distribution, respectively. `AIC` means Akaike information criterion.
\end{tablenotes}
\end{threeparttable}}
\end{table}

Figure \ref{fig:figure-03} illustrates these fitted functions (red line)
alongside histograms of empirical travel times (grey bars). As for the
others pre-estimated functions, these examples allow to calculate
gravity-based accessibility measures for active travel across various
destinations and temporal scales in Canadian urban areas.

{\centering \includegraphics[width=1\linewidth]{Manuscript-figures/impf_Work or school}

\caption{Empirical data and impedance functions fitted for `work or school` as destination.}\label{fig:figure-03}
\end{figure}

\hypertarget{concluding-remarks}{%
\section{Concluding remarks}\label{concluding-remarks}}

This article presents \texttt{ActiveCA}, an open data product that
provides analysis-ready data from Cycles 2 (1986), 7 (1992), 12 (1998),
19 (2005), 24 (2010), and 29 (2015) of the GSS surveys on active travel
in Canada. In the form of an \texttt{R} data package, \texttt{ActiveCA}
was developed after collecting, cleaning, and processing the survey
data, providing information on origins, destinations, and duration of
active travel, as well other information. Additionally, the package
includes a series of pre-estimated impedance functions for walking and
cycling trips, considering various destinations and time periods, for
Canadian Metropolitan and Census Agglomeration areas.

The value of \texttt{ActiveCA} lies in its transparency, accessibility,
and open infrastructure, which facilitates the addition of complementary
data sets in the future. R users can seamlessly explore GSS walking and
cycling episodes along with calibrated impedance functions, with the
option to suggest enhancements to the package as needed. This article
adopts the structure proposed by Anastasia and Páez
\citeyearpar{soukhov2023}, whose work provided essential guidance for
the creation of this package. Similarly, we aim to contribute to the
academic community by promoting transparent research practices that
encourage replication and innovation in related fields. We believe that
\texttt{ActiveCA} will serve as a basis for further research on GSS and
for the integration of additional data by the authors or the wider open
source community.

\hypertarget{declaration-of-conflicting-interests}{%
\section{Declaration of Conflicting
Interests}\label{declaration-of-conflicting-interests}}

@@

\hypertarget{data-availability-statement}{%
\section{Data availability
statement}\label{data-availability-statement}}

The \{ActiceCA\} R data package can be found and installed on Github
(\emph{link}).

>>>>

\begin{abstract}
This paper describes \{ActiveCA\}, and open data product with Canadian
time use data. \{ActiveCA\} is an \texttt{R} data package that contains
analysis-ready data related to active travel spanning almost 40 years,
extracted from cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24
(2010), and 29 (2015) of Canada's General Social Survey. Active travel
is characterized by mode, with walking being part of every cycle and
bicycling starting in 1992. The attributes of active trips are the types
of locations of origins and destinations, the duration of trips, and
episode weights for expanding the trips to population-wide estimates.
Based on the year of the survey, a variety of locations are coded. In
earlier cycles, these include home, work or school, and other's home,
whereas in later cycles these are augmented with locations such as
grocery stores, restaurants, outdoor destinations, and others. The
geographical resolution includes the province and whether the episode
was in an urban or rural setting.
\end{abstract}

\keywords{Active; mobility; walking; cycling; travel time; time-use;
General Social Survey}

@@

\section{Introduction}\label{introduction}

The objective of this paper is to introduce \{ActiveCA\}, an open data
product with time use data from Canada's General Social Surveys. Open
data products (ODPs) are the outcome of a process that transforms raw
data (open or not) into analysis-ready data, following a transparent
process in which all stages of development follow open principles
\citep{arribas-bel2021}. ODPs, while still open, differ from general
open data in their degree of ease of access, their heightened usability,
and potentially the value they add to the raw data.

\{ActiveCA\} provides analysis-ready data concerning active travel in
Canada spanning a period of almost 40 years. The source of these data is
Canada's program of General Social Surveys (GSSs). This program is
designed to provide cross-sectional data on topics of interest to
improve the well-being of Canadians. As part of this program, every five
to seven years the survey is done on the topic of time use. Concretely,
\{ActiveCA\} covers Cycles 2 (1986), 7 (1992), 12 (1998), 19 (2005), 24
(2010), and 29 (2015) of the GSS. Time use data in these surveys is
coded using a very fine grain, from time spent in chores, leisure, and
sleeping, to time spent working or at school. These surveys have proved
valuable in investigations of mobility and quality of life
\citep{spinneyTransport2009}, the relationship between active travel and
transit use \citep{lachapelleLonger2016}, and travel behavior and time
poverty \citep{kimFacing2024}, to name but a few examples.

For \{ActiveCA\}, data we extracted all data needed to characterize
active travel in Canada, namely, episodes where the activity was
identified as moving between an origin and a destination, either by
walking or cycling. Although Statistics Canada offers Public Use
Microdata Files and documentation for the GSS program
\citep[see][]{statisticscanada2024}, accessing these files, and
preparing them for analysis is not a straightforward matter, given their
size and complexity. The process of extracting information of interest
from the source files is time-consuming, tedious, and challenging and/or
prone to error due to the expertise required to work with these files.
To create \{ActiveCA\} we collected, cleaned, and processed the cycles
of the GSS surveys concerning time use to make them ready for analysis.

\{ActiveCA\} is distributed as an \texttt{R} package with a number of
data objects and their documentation. \texttt{R} packages contain code,
data, and documentation in a standardized format that can be installed
by \texttt{R} users via a software repository, such as CRAN
(Comprehensive R Archive Network) or GitHub, which makes them an adroit
medium to distribute analysis-ready data.

Given the level of interest in active travel
\citep[e.g.,][]{mccurdySupport2023}, reducing the barriers to using data
contained in rich, but difficult to access and use surveys, such as
Canada's GSSs, is a worthy endeavour that can only improve data-driven
decisions in transportation, urban, and health policy.

The rest of this paper discusses the sources of data, and the process
implemented to retrieve and package them. Then, we explain how to use
the package and show some selected examples of analysis to whet the
imagination of potential users. This ODP provides not only data that are
easy to use, but also all the code and documentation that make this a
reproducible research project. In summary, \{ActiveCA\} aims to
implement and inspire the best principles of open spatial sciences
\citep{paez_open_2021, brunsdon_opening_2021}.

\section{General Social Survey (GSS)
collection}\label{general-social-survey-gss-collection}

Statistics Canada \citeyearpar{statisticscanada2024} conducts GSS
surveys to obtain data on social trends to track changes in Canadians'
living conditions and well-being over time. The series of survey on time
use are used to understand how Canadian residents spend and manage their
time, and what factors contribute to their happiness and stress. The GSS
program was created in 1985, and is serialized to provide a collection
of annual, representative cross-sectional surveys.

The topics of the survey cycle every few years to cover topics that
include family, health, social identity, and every five to seven years
time use. The first Canadian time use survey done as part of the GSS
program was conducted in 1986, and the most recent was completed in
2015. These Time Use Surveys \citep{statisticscanada2022} collect data
on respondents' participation and time spent on a wide range of everyday
activities using a 24-hour retrospective diary, with information on the
location of these activities (e.g.~at home, at work, etc.) and, for
non-personal activities, the people who were present with the respondent
at the time of the activity. In addition, time-use surveys also cover
topics related to leisure time, work-life balance, health, commuting,
culture and sports, and many others.

The Public Use Microdata are released by Statistics Canada in two files:
a \emph{Main File} and an \emph{Episode File}. The files are linked by
keys that identify households, individuals, and episodes (i.e.,
activities) conducted by individuals. We discuss these files in more
detail in the following section.

\subsection{The Main File}\label{the-main-file}

The main file of the time use survey compiles a large array of
aggregated data, summarizing the answers to the questionnaire that
describe households and individuals, as well as derived variables that
summarize the respondents' use of time use across different activities,
locations, and social interactions. This file documents the time and
duration that respondents allocate to each activity and location. The
Main File provides a overview of daily routines and social dynamics, not
focusing on individual activity episodes. Additionally, this file
categorizes activities into bigger groups and subcategories,
facilitating the data's analytical utility with additional metrics such
as total transit time, time spent with household members, and counts of
activities and episodes.

\subsection{The Episode File}\label{the-episode-file}

The episode is a much bigger file that records detailed data for each
activity episode reported by respondents. Each episode represents a
single activity and its duration, and the sum of all episodes throughout
the day adds up to 24 hours. Each entry in this file includes the start
and end times of the activity, the duration, location, and accompanying
social context, informing when and where activities occurred and with
whom. The focus of the Episode File is not on the characteristics of the
respondents, but on the characteristics of the activities, and the data
are structured around the numerous activity instances that compose a day
of the respondent. Although respondent-specific characteristics are not
included within the episode file, it is possible to link the main file
and the episode file by using a key present in both the Main and Episode
Files.

\section{Methods}\label{methods}

For each selected cycle of the GSS surveys, we reviewed the episode
files to identify episodes of movement that involved walking or cycling.
This allowed us to select the activities immediately before and after
the movement episode. After that, we labeled the code variables with
their appropriate descriptions, identifying each origin and destination,
mode of travel, as well the province and urban classification of the
episode.

\section{How to use \{ActiveCA\}}\label{how-to-use-activeca}

\subsection{Descriptive statistics}\label{descriptive-statistics}

@@

\{ActiveCA\} also enables visual analysis of active travel in Canada
using traditional exploratory data analysis techniques. Figures
\ref{fig:figure-01} and \ref{fig:figure-02} show walking and cycling
trips from 1992 and 2015 through heat maps. These maps use color
gradients to represent the percentage of trips between various origins
and destinations, with darker colors indicating higher percentages and
lighter colors representing less frequent routes. For conciseness, we
omitted the heat maps for the other years analyzed.

@@

\section{Python integration}\label{python-integration}

CAN WE PROVIDE PYTHON INTEGRATION? See here:

\url{https://github.com/paezha/idealista18/tree/master/Python}

\section{Concluding remarks}\label{concluding-remarks}

This paper presents \texttt{ActiveCA}, an open data product that
provides analysis-ready data from Cycles 2 (1986), 7 (1992), 12 (1998),
19 (2005), 24 (2010), and 29 (2015) of the GSS surveys on active travel
in Canada. In the form of an \texttt{R} data package, \{ActiveCA\} was
developed after collecting, cleaning, and processing the survey data,
providing information on origins, destinations, and duration of active
travel, as well other information.

It is important to remark that the present version of \{ActiveCA\}
covers all Canadian time use surveys up to 2015. While the most recent
time use survey was carried out in 2022 \citep{wray2024}, the Public Use
Microdata Files are currently unavailable, and at the moment it is
estimated that they will only be published in the later part of 2025.
The \texttt{R} package will be updated once these files become
available.

The value of \{ActiveCA\} lies in its transparency, accessibility, and
ease of use, which facilitates the addition of complementary data sets
in the future. \texttt{R} users can seamlessly explore GSS walking and
cycling episodes , with the option to suggest enhancements to the
package as needed. This article adopts the structure proposed by
Anastasia and Páez \citeyearpar{soukhov2023}, whose work provided
essential guidance for the creation of this package. Similarly, we aim
to contribute to the academic community by promoting transparent
research practices that encourage replication and innovation in related
fields. We believe that \{ActiveCA\} will serve as a basis for further
research on GSS and for the integration of additional data by the
authors or the wider open source community.

\section{Declaration of Conflicting
Interests}\label{declaration-of-conflicting-interests}

@@

\section{Data availability statement}\label{data-availability-statement}

The \{ActiveCA\} R data package can be found and installed on Github
(\emph{link}).
